Discrete Probability for Discrete Quantum Computing

### 2016-05-05 prob.tex
prob.tex \section{Classical Probability Spaces} something like this?

Instead of using every point in the real interval $[0,1]$ we can
partition the interval into disjoint sets and only consider
probability measures up to set membership. The simplest such situation
is to partition the interval $[0,1]$ into $\{0\}$ (which we will call
$\imposs$) and the half-open interval $(0,1]$ (which we will call
$\poss$). The addition that was used to aggregate probabilities is now
abstracted to $\vee$ such that $x \vee y=\imposs$ if and only if
$x=y=\imposs$. We will call the resulting set
$\left\{ \imposs, \poss\right\}$ together with the with associated
operation $\vee$, the set $\mathscr{L}_{2}$. The definition of a
probability measure in this case is modified to a function
$\pmeas : \events \rightarrow \mathscr{L}_{2}$ such that:
\begin{itemize}
\item $\pmeas(\Omega) = \poss$, and
\item for a collection $E_i$ of pairwise disjoint events, $\pmeas(\bigcup_i
E_i) = \bigvee_i \pmeas(E_i)$.
\end{itemize}

\begin{example}[Two-coin probability space with finite set-valued
  probability measure] \label{ex2} Under the new set-valued
  requirement, the probability measure in the first example becomes:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) &=& \imposs \\
\pmeas(\{ HH \}) &=& \poss \\
\pmeas(\{ HT \}) &=& \imposs \\
\pmeas(\{ TH \}) &=& \poss \\
\pmeas(\{ TT \}) &=& \imposs \\
\pmeas(\{  HH, HT \}) &=& \poss \\
\pmeas(\{  HH, TH \}) &=& \poss \\
\pmeas(\{  HH , TT \}) &=& \poss \\
\end{array} & \begin{array}{rcl}
\pmeas(\{  HT, TH \}) &=& \poss \\
\pmeas(\{  HT , TT \}) &=& \imposs \\
\pmeas(\{  TH , TT \}) &=& \poss \\
\pmeas(\{  HH, HT, TH \}) &=& \poss \\
\pmeas(\{  HH, HT, TT \}) &=& \poss \\
\pmeas(\{  HH, TH, TT \}) &=& \poss \\
\pmeas(\{  HT, TH, TT \}) &=& \poss \\
\pmeas(\{  HH, HT, TH, TT \}) &=& \poss 
\end{array}
\end{array}\]
Despite the fact that we have lost all numeric information, the
probability measure still reveals that the second coin is
double-headed. We have however lost the information regarding the bias
in the first coin. This information can be recovered with a more
refined probability measure as we show next.  
\qed\end{example}

%%%%%
\subsection{Interval-valued probability measures}

A natural generalization of the disjoint set-valued measure above is
to allow the sets to overlap. In this case, we split the interval
$[0,1]$ in a collection of \emph{overlapping} closed
sub-intervals. First we illustrate the main ideas using a simple
example.

\begin{example}[Two-coin probability space with four
  intervals] \label{ex3} We split the unit interval $[0,1]$ in the
  following four closed sub-intervals: $[0,0]$ which we call \imposs,
  $[0,\frac{1}{2}]$ which we call \unlikely, $[\frac{1}{2},1]$ which
  we call \likely, and $[1,1]$ which we call \necess.  Using these new
  values, we can modify the probability measure of Ex.~\ref{ex1} by
  mapping each numeric value to the smallest sub-interval containing
  it to get the following:
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \necess\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \necess\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \necess\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \necess
\end{array}\end{array}
\]
This probability measure is more informative than the one in
Ex.~\ref{ex2}: not only does it reveal that the second coin is
double-headed but it also reveals the bias in the first coin.
\qed\end{example}

The probability measure above appears quite intuitive but it is not
really evident that it is well-defined. For example, how do we justify
the following combination of assignments:
\[ 
\pmeas(\{HH\}) = \unlikely, \quad\pmeas(\{TH\}) = \likely, \quad
\pmeas(\{HH\}\cup\{TH\}) = \necess 
\]
which assert that an \unlikely-event whose probability is in the range
$[0,\frac{1}{2}]$ and a \likely-event whose probability is in the
range $[\frac{1}{2},1]$ combine to a \necess-event whose probability
is in the range $[1,1]$. To understand the calculation, we provide the
formal definition of a probability measure in this case.

Fix a collection $\mathscr{I}$ of closed sub-intervals of $[0,1]$ that
must include $[0,0]$ and $[1,1]$. An $\mathscr{I}$--interval-valued
probability measure is a function
$\mu : \events \rightarrow \mathscr{I}$ such that:
\begin{itemize}
\item $\mu(\emptyset) = [0,0]$, 
\item $\mu(\Omega) = [1,1]$, and 
\item for any mutually disjoint events $E_1$, $E_2$, $E_3$, and $E_4$
  where $\pmeas(E_i) = [l_i,h_i]$ and where
  $E_1 \cup E_2 \cup E_3 \cup E_4 = \Omega$, we have:
\[
\pmeas(\{E_1\cup E_2\}) \subseteq
  [\max(l_1+l_2 , 1-(h_3+h_4)), \min(h_1+h_2 , 1-(l_3+l_4))]
\]
We view the events $E_1$ and $E_2$ as providing evidence for the
combined event $E_1\cup E_2$: their values set positive bounds on the
resulting interval. But as $E_3 \cup E_4$ is the complement of
$E_1 \cup E_2$, the events $E_3$ and $E_4$ provide negative bounds on
the resulting interval. The calculated interval may not be in our
fixed set $\mathscr{I}$ of chosen intervals so we embed it in the
smallest existing interval.
\end{itemize}

>>>>

Instead of using every point in the real interval $[0,1]$ we can
consider subsets~$\mathscr{L}$ of $[0,1]$ and only consider probability
measures up to set membership with intuitive addition\footnote{Notice that the set-valued measure in~\cite{Artstein1972,PuriRalescu1983},
only require $I_{1}+I_{2}=\left\{ a_{1}+a_{2}\middle|a_{1}\in I_{1},a_{2}\in I_{2}\right\} $
because they focused on measure theory. In probability thoery, $2\in\pmeas\left(E\right)$
doesn't really make sense so that we consider intersect $[0,1]$ in
our definition.} 
\[
I_{1}+I_{2}=\left\{ a_{1}+a_{2}\middle|a_{1}\in I_{1},a_{2}\in I_{2}\right\} \cap[0,1]\textrm{ .}
\]
Then, a probability measure in this case is modified to a function
$\pmeas:\events\rightarrow\mathscr{L}$ such that: 
\begin{itemize}
\item $1\in\pmeas(\Omega)$, and 
\item for a collection $E_{i}$ of pairwise disjoint events, $\pmeas(\bigcup_{i}E_{i})=\sum_{i}\pmeas(E_{i})$. 
\end{itemize}
The simplest such situation is to partition the interval $[0,1]$
into $\{0\}$ (which we will call $\imposs$) and the half-open interval
$(0,1]$ (which we will call $\poss$). We will call the resulting
set $\left\{ \imposs,\poss\right\} $, the set $\mathscr{L}_{2}$.

\begin{example}{[}Two-coin probability space with finite set-valued
probability measure{]} \label{ex2} Under the new set-valued requirement,
the probability measure in the first example becomes: 
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \poss\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \poss\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \poss\\
\pmeas(\{HH,TH\}) & = & \poss\\
\pmeas(\{HH,TT\}) & = & \poss
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \poss\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \poss\\
\pmeas(\{HH,HT,TH\}) & = & \poss\\
\pmeas(\{HH,HT,TT\}) & = & \poss\\
\pmeas(\{HH,TH,TT\}) & = & \poss\\
\pmeas(\{HT,TH,TT\}) & = & \poss\\
\pmeas(\{HH,HT,TH,TT\}) & = & \poss
\end{array}\end{array}
\]
Despite the fact that we have lost all numeric information, the probability
measure still reveals that the second coin is double-headed. We have
however lost the information regarding the bias in the first coin.
This information can be recovered with a more refined probability
measure as we show next. \qed\end{example}

Although $\mathscr{L}_{2}$-valued probability measure is quite intuitive,
set-valued probability measure is not that intuitive if we extend
to more values. For example, consider the following three \emph{overlapping}
closed sub-intervals: $[0,0]$ is still call \imposs, $[0,\frac{1}{2}]$
which we call \unlikely, ad $[\frac{1}{2},1]$ which we call \likely.
The following example is a set-valued probability measure corresponding
to the probability measure in the first example.

\begin{example}\label{Two-coin-set-valued-again}{[}Two-coin probability
space with set-valued probability measure, again{]} 
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \likely\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \likely\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \likely\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \likely
\end{array}\end{array}
\]
In this example, we can get the information that the first coin is
weighted and the second coin is double-headed. However, we may notice
that the set-valued probability of the whole space~$\{HH,HT,TH,TT\}$
are different between these two examples. This problem can be fixed
when we moved on to the next section. \qed\end{example}

We will return to finite set-valued probability measures in Sec.~\ref{sec:?}.

%%%%%

\subsection{Interval-valued probability measures}

The singleton closed interval $[0,0]$ is a required spacial value
for empty set, and a natural generalization is to consider another
special closed interval $[1,1]$ which we call \necess for the whole
space. In general, if the probability of an event $E$ is $[a,b]$,
we think of the left-endpoint~$a$ as representing the strength of
the evidence that supports $E$, and the right-endpoint~$b$ as the
strength of the evidence that contradicts $E$.

Thus if we have an event $E$ with probability $[a,b]$ where $a=0.1$
and $b=0.7$, we have that: 
\begin{itemize}
\item the strength of evidence supporting $E$ is 0.1; since either $E$
or its complement must happen, we conclude that there is 0.9 evidence
supporting the complement of $E$; 
\item the strength of evidence contradicting $E$ is 0.7; again since either
$E$ or its complement must happen, we conclude that there is 0.3
evidence contradicting the complement of $E$. 
\end{itemize}
\yutsung{Do we use the law of excluded middle here? You remind me
Agda : ) Did Homotopy Type Theory people said anything about the probability?}\\
Turning things around, the strength of evidence that contradicts $E$
is evidence supporting the complement of $E$. The complement of $E$
must therefore have probability $\left[1-b,1-a\right]$ which we abbreviate
$1-\left[a,b\right]$, so the probability measure $\pmeas$ should
satisfy:
\begin{itemize}
\item $\pmeas(\emptyset)=[0,0]$, 
\item $\pmeas(\Omega)=[1,1]$, and 
\item $\pmeas\left(\Omega\backslash E\right)=1-\pmeas\left(E\right)$
\end{itemize}
However, if we want $\pmeas(\Omega)=[1,1]$, and the probability assignment
for singlton sets $\pmeas(\{HH\})=\unlikely$, $\pmeas(\{HT\})=\imposs$,
$\pmeas(\{TH\})=\likely$, and $\pmeas(\{TT\})=\imposs$ as in example~\ref{Two-coin-set-valued-again},
we only have $\pmeas(\Omega)\subsetneq\pmeas(\{HH\})+\pmeas(\{HT\})+\pmeas(\{TH\})+\pmeas(\{TT\})$.
In general, we only require
\begin{itemize}
\item for a collection $E_{i}$ of pairwise disjoint events, we have $\pmeas\left(\bigcup_{i}E_{i}\right)\subseteq\sum_{i}\pmeas\left(E_{i}\right)$. 
\end{itemize}
This statement means that the evidences of $\bigcup_{i}E_{i}$ is
at least as strong as putting all the evidences of $E_{i}$ together,
but some evidence may only be acquired for $\bigcup_{i}E_{i}$ as
the whole. Therefore, $\pmeas\left(\bigcup_{i}E_{i}\right)$ is a
subset of $\sum_{i}\pmeas\left(E_{i}\right)$, but may not equal.
In our example, 
\begin{eqnarray*}
 &  & \pmeas(\{HH\})+\pmeas(\{HT\})+\pmeas(\{TH\})+\pmeas(\{TT\})\\
 & = & \imposs+\unlikely+\imposs+\likely\\
 & = & \left[0,0\right]+\left[0,\frac{1}{2}\right]+\left[0,0\right]+\left[\frac{1}{2},1\right]=\left[\frac{1}{2},\frac{3}{2}\right]
\end{eqnarray*}
The above equation told us $\pmeas(\Omega)\subseteq\left[\frac{1}{2},\frac{3}{2}\right]$.
However, because of $\pmeas(\emptyset)=\imposs$, we have $\pmeas\left(\Omega\right)=1-\pmeas\left(\emptyset\right)=\necess$,
but $\pmeas(\emptyset)=\imposs$ cannot be used to reasoning the probability
of $\{HH\}$, $\{HT\}$, $\{TH\}$ and $\{TT\}$ individually. The
full probability assignment is shown in the following example.

\begin{example}{[}Two-coin probability space with four intervals{]}
\label{ex3} We split the unit interval $[0,1]$ in the following
four closed sub-intervals: $[0,0]$ which we call \imposs, $[0,\frac{1}{2}]$
which we call \unlikely, $[\frac{1}{2},1]$ which we call \likely,
and $[1,1]$ which we call \necess. Using these new values, we can
modify the probability measure of Ex.~\ref{ex1} by mapping each
numeric value to the smallest sub-interval containing it to get the
following: 
\[
\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) & = & \imposs\\
\pmeas(\{HH\}) & = & \unlikely\\
\pmeas(\{HT\}) & = & \imposs\\
\pmeas(\{TH\}) & = & \likely\\
\pmeas(\{TT\}) & = & \imposs\\
\pmeas(\{HH,HT\}) & = & \unlikely\\
\pmeas(\{HH,TH\}) & = & \necess\\
\pmeas(\{HH,TT\}) & = & \unlikely
\end{array} & \begin{array}{rcl}
\pmeas(\{HT,TH\}) & = & \likely\\
\pmeas(\{HT,TT\}) & = & \imposs\\
\pmeas(\{TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH\}) & = & \necess\\
\pmeas(\{HH,HT,TT\}) & = & \unlikely\\
\pmeas(\{HH,TH,TT\}) & = & \necess\\
\pmeas(\{HT,TH,TT\}) & = & \likely\\
\pmeas(\{HH,HT,TH,TT\}) & = & \necess
\end{array}\end{array}
\]
This probability measure is more informative than the one in Ex.~\ref{ex2}:
not only does it reveal that the second coin is double-headed but
it also reveals the bias in the first coin. \qed\end{example}
